\documentclass[a4paper]{report}

\makeatletter\@addtoreset{chapter}{part}\makeatother%

\usepackage[margin=2cm]{geometry}
\usepackage{parskip}

\begin{document}

\begin{titlepage}
	\raggedleft
	\vspace*{2cm}
	\Huge{Computer Science Notes}\\
	\vspace{0.5cm}
	\large{OCR H446}
\end{titlepage}

\tableofcontents

\part{Computer systems}

\chapter{Components of a computer and their uses}

\section{Structure and function of the CPU}

The CPU is made up of 3 main components:

\begin{itemize}
	\item{Units}
	\item{Registers}
	\item{Buses}
\end{itemize}

\subsection{Units}

Units: Performs operations within the CPU.

2 main units:

\begin{itemize}
	\item{Control Unit (CU)}
	\item{Arithmetic and Logic Unit (ALU)}
\end{itemize}

\subsubsection{Control Unit}

CU: directs the operation of the CPU. Its resposibilities are as follows:

\begin{itemize}
	\item{Managing and controlling the activities the CPU performs.}
	\item{Managing and controlling the flow of data between the CPU and
		other components.}
	\item{Accepting and decoding instructions.}
	\item{Storing the results of data back into main memory.}
\end{itemize}

\subsubsection{Arithmetic and Logic Unit}

ALU: performs arithmetic and logical operations on data. Its capabilities are
as follows:

\begin{itemize}
	\item{Arithmetic operations:}
		\begin{itemize}
			\item{Addition}
			\item{Subtraction}
		\end{itemize}
	\item{Boolean logic operations:}
		\begin{itemize}
			\item{AND}
			\item{OR}
			\item{XOR}
			\item{NOT}
		\end{itemize}
\end{itemize}

The result of any calculations performed by the ALU is stored in the
Accumulator register (ACC).

\subsection{Registers}

Registers: small, extremely fast memory cells that are available to the CPU.
Used to temporarily store any data that is being worked on by the CPU (this
could be an instruction, an address, or some other value).

5 main registers:

\begin{itemize}
	\item{Program Counter (PC): holds the address of the next instruction
		be executed.}
	\item{Accumulator (ACC): holds the result of any calculation done by
		the ALU.}
	\item{Memory Address Register (MAR): holds the address of a location
		to be read from/written to.}
	\item{Memory Data Register (MDR): holds data that has been read/is to
		be written.}
	\item{Current Instruction Register (CIR): holds the current
		instruction being executed.}
\end{itemize}

\subsection{Buses}

Bus: connects the CPU to other parts of the computer (cheifly main memory).

Width of a bus: the number of lanes that a bus has. Proportional to the number
of bits that the bus can transmit simultaneously at any given moment.

3 main buses:

\begin{itemize}
	\item{Address bus}
	\item{Control bus}
	\item{Data bus}
\end{itemize}

\subsubsection{Address bus}

Address bus: used for transfer of memory addresses where data is to be read
from/written to. Is unidirectional.

The CPU transmits addresses it needs to access across this bus to main memory,
which can then either send the data at that address back along the data bus,
or can open that address for writing.

Width of this bus is proportional to the number of addressable locations -
a 32 bit width address bus can address 4294967296 (2\^32) discrete locations
in main memory.

\subsubsection{Control Bus}

Control bus: used for transfer of control signals between the components of
the computer, and controls the use of the other buses. Is bidirectional.

Can transmit a variety of signals:

\begin{itemize}
	\item{Bus request: component wants to access data bus.}
	\item{Bus grant: component granted access to data bus.}
	\item{Interrupt request: component requires immediate attention from
		the CPU.}
	\item{Memory read: data from an address is to be placed onto data
		bus.}
	\item{Memory write: data on data bus is to be written to address.}
	\item{Clock: synchronise operations.}
\end{itemize}

\subsubsection{Data Bus}

Data bus: used for transfer of data and instructions. Is bidirectional.

\subsection{Fetch-Decode-Execute cycle}

Fetch-decode-execute cycle: process carried out by the CPU in order to execute
an instruction. Follows the following steps:

\begin{enumerate}
	\item{Fetch}
		\begin{enumerate}
			\item{Address in PC is copied to MAR}
			\item{Address in MAR is requested from RAM via address
				bus}
			\item{Instruction at address is copied to MDR via data
				bus}
			\item{PC is incremented by 1}
			\item{Instruction in MDR is copied to CIR}
		\end{enumerate}
	\item{Decode}
		\begin{enumerate}
			\item{Instruction in CIR is decoded by CU (split into
				opcode and operand - see below for more
				details)}
		\end{enumerate}
	\item{Execute}
		\begin{enumerate}
			\item{Instruction in CIR is executed}
			\item{Check for and handle any interrupts}
		\end{enumerate}
\end{enumerate}

\subsubsection{Instruction structure}

Opcode: type of instruction to be executed.

Operand: either data to be operated on or address which contains data to be
operated on.

\chapter{Software and software development}

\chapter{Exchanging data}

\chapter{Data types, data structures and algorithms}

\chapter{Legal, moral, cultural and ethical issues}

\part{Algoriths and programming}

\chapter{Elements of computational thinking}

\chapter{Problem solving and programming}

\chapter{Algorithms}

\end{document}
