\documentclass[a4paper]{report}

\makeatletter\@addtoreset{chapter}{part}\makeatother%

\usepackage[margin=2cm]{geometry}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage{amsfonts}

\begin{document}

\begin{titlepage}
	\raggedleft
	\vspace*{2cm}
	\Huge{Mathematics Notes}\\
	\vspace{0.5cm}
	\large{OCR H240}
\end{titlepage}

\tableofcontents

\part{Pure Mathematics}

\chapter{Proof}

\section{Mathematical structures and notation}

Basic mathematical structures:

\begin{itemize}
	\item{Statement/expression: most basic structure. Example: \( x + 3 \).}
	\item{Equation: statements involving equals sign (\( = \)). Only true for discrete values
		of unknown(s). Example: \( x^2 + 3x + 2 = 0 \).}
	\item{Inequality: statements involving inequality sign (\( <, >, \leq, \geq \)). Only true
		for a range of values of the unknown(s). Example: \( 3x + 7 > 4 \).}
	\item{Identity/congruence: statements involving an equivalence sign (\( \equiv, \iff \)).
		True for all values of unknown variable(s). Example:
		\( x^3 - x \equiv x(x + 1)(x - 1) \).}
\end{itemize}

Implication:

\begin{itemize}
	\item{\( A \implies B \): if \( A \) is true then \( B \) is true/\( A \) is sufficient
		for \( B \).}
	\item{\( A \impliedby B \): if \( B \) is true then \( A \) is true/\( A \) is necessary
		for \( B \).}
\end{itemize}

Important sets:

\begin{itemize}
	\item{\( \emptyset \): null set.}
	\item{\( \mathbb{N} \): set of natural numbers (\( 1, 2, 3, \dots \)).}
	\item{\( \mathbb{Z} \): set of integers (\(0, \pm1, \pm2, \pm3, \dots \)).}
	\item{\( \mathbb{R} \):  set of real numbers.} 
\end{itemize}

Set notation:

\begin{itemize}
	\item{\( x > 9 \) becomes \( \{ x : x > 9 \} \).}
	\item{\( x \leq 11 \) becomes \( \{ x : x \leq 11 \} \).}
	\item{\( 3 < x \leq 5 \) becomes \( \{ x : x > 3 \} \cap \{ x : x \leq 5 \} \).}
	\item{\( x > 7 \) or \( x < 2 \) becomes \( \{ x : x > 7 \} \cup \{ x : x < 2 \} \).}
\end{itemize}

Interval notation:

\begin{itemize}
	\item{\( x > 9 \) becomes \( x \in (9, \infty) \).}
	\item{\( x \leq 11 \) becomes \( x \in (-\infty, 11] \).}
	\item{\( 3 < x \leq 5 \) becomes \( x \in (3, 5] \).}
	\item{\( x > 7 \) or \( x < 2 \) becomes \( x \in (7, \infty) \cup (-\infty, 2) \).}
\end{itemize}

Set and interval notation used to show solutions for inequalities. Assumed that
\( x \in \mathbb{R} \) unless otherwise.

\section{Disproof by counterexample}

Disproving a hypothesis by finding an example that contradicts it. Example: \[ \text{Show that }
\sqrt{x^2} \text{ is not always equal to }  x \]

Let \( x = -4 \). Then: \[ x^2 = 16, \sqrt{x^2} = \pm 4 \]

\( x \neq \pm 4 \) therefore the hypothesis is incorrect.

\section{Proof by deduction}

Proving a hypothesis using algebra and logic. Example: \[ \text{Show that the sum of 3 consecutive
numbers is a multiple of 3} \]

Let 3 consecutive numbers be \( n, n + 1, n + 2 \). Then: \[ n + n + 1 + n + 2 = 3n + 3 \]

\( 3n + 3 \) can be factorised to \( 3(n + 1) \), which is a multiple of 3, therefore the
hypothesis has been proved.

\section{Proof by exhaustion}

Proving a hypothesis by checking every possible case. Example: \[ \text{Prove that no square number
less than 100 ends in 8} \]

The square numbers less than 100 are: 1, 4, 9, 16, 25, 36, 49, 64, 81, 100. None of these end in 7
therefore the hypothesis has been proved.

\section{Proof by contradiction}

Proving a hypothesis by trying to prove the opposite of the hypothesis and showing that this leads
to an invalid conclusion. Example: \[ \text{Prove that } \sqrt{2} \text{ is irrational} \]

Let \( \sqrt{2} = \frac{a}{b} \) such that \( a, b \in \mathbb{Z}, \gcd(a, b) = 1 \). Then: \[ 2 =
\frac{a^2}{b^2}, 2b^2 = a^2 \]

Therefore \( a^2 \) is a multiple of 2, therefore \( a \) is a multiple of 2. Let \( a = 2k \).
Then \[ 2b^2 = (2k)^2, 2b^2 = 4k^2, b^2 = 2k^2 \]

Therefore \( b^2 \) is a multiple of 2, therefore \( b \) is a multiple of 2. This contradicts
\( \gcd(a, b) = 1 \), therefore \( \sqrt{2} \) is irrational.

Another example: \[ \text{Prove that there are an infinite number of primes} \]

Assume that there are a finite number (\( n \)) of primes: \( p_1, p_2, \dots, p_n \). Let \( P =
p_1 \times p_2 \times \dots p_n + 1 \). \( P \) is not a prime as it is greater than any of the
primes, but \( P \) is not divisible by any of the primes. Therefore either \( P \) is prime or
\( P \) is divisible by a prime greater than \( p_n \). Either way, we have proved the hypothesis.

\chapter{Algebra and Functions}

\chapter{Coordinate Geometry in the \( x - y \) plane}

\chapter{Sequences and Series}

\chapter{Trigonometry}

\chapter{Exponentials and Logarithms}

\chapter{Differentiation}

\chapter{Integration}

\chapter{Numerical Methods}

\chapter{Vectors}

\part{Statistics}

\chapter{Statistical Sampling}

\chapter{Data Representation and Interpretation}

\chapter{Probability}

\chapter{Statistical Distributions}

\chapter{Statistical Hypothesis Testing}

\part{Mechanics}

\chapter{Quantities and Units in Mechanics}

\chapter{Kinematics}

\chapter{Forces and Newton's Laws}

\chapter{Moments}

\end{document}
