\documentclass[a4paper]{report}

\makeatletter
\renewcommand{\@chapapp}{Week}
\makeatother

\usepackage[margin=2cm]{geometry}
\usepackage{parskip}

\usepackage{tikz}

\usetikzlibrary{arrows, shapes.geometric, positioning}
\tikzset{node distance=1cm, auto}
\tikzstyle{process} = [rectangle, rounded corners, minimum width=3cm,
minimum height=1cm, text centered, draw=black]
\tikzstyle{decision} = [diamond, rounded corners, minimum width=3cm,
minimum height=3cm, text centered, draw=black]
\tikzstyle{goal} = [draw, ellipse, minimum width=3cm, minimum height=1cm,
draw=black]
\tikzstyle{line} = [draw, -latex']

\usepackage{listings}

\lstset{
	basicstyle=\ttfamily,
}
\lstdefinelanguage{pseudo}{
	morekeywords={while, if, else, for, function, return, start, stop},
	sensitive=false,
	morecomment=[l]{\#},
	escapechar=\%
}

\begin{document}

\begin{titlepage}
	\raggedleft
	\vspace*{2cm}
	\Huge{Programming Notes}\\
	\vspace{0.5cm}
	\large{ECM1400}
\end{titlepage}

\tableofcontents

\chapter{Problem solving with code}

\section{Notes}

\subsection{Programming concepts}

Programming: process of writing computer code to solve real world problems.
Expressed by the following:

\begin{itemize}
	\item{Being familiar with basic constructs/concepts of a programming
		language.}
	\item{Breaking a problem down into smaller problems that are solved
		using computer programs.}
\end{itemize}

Programming language: a formal language used to dictate the actions of a
computer.

Program: a set of instructions written in a programming language.

Algorithm: a well defined, finite process for achieving a certain outcome.
Examples include an algorithm to sort a list of numbers. Basic elements of
algorithms:

\begin{itemize}
	\item{Loops: repitition of an action.}
		\begin{itemize}
			\item{Repeating till a condition is reached.}
			\item{Repeating a specific number of times.}
		\end{itemize}
	\item{Conditionals/branching: change behaviour depending on input.}
	\item{Functions: encapsulation of parts of the algorithm.}
	\item{Variables: data the algorithm works on.}
\end{itemize}

Representations of algorithms:

\begin{itemize}
	\item{Visual depiction through flowcharts.}
	\item{Natural language description through pseudocode.}
\end{itemize}

\subsection{Flowcharts}

Flowchart: a diagram which represents an algorithm. Basic elements of flowcharts:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\node (process) [process] {Process};
		\node (function) [process, below=of process]
		{\underline{Function}};
		\node (start) [goal, right=of process] {Start};
		\node (stop) [goal, below=of start] {Stop};
		\node (decision) [decision, right=of start] {Decision};
	\end{tikzpicture}
\end{figure}

Flowcharts begin with the \textit{Start} element and end with the
\textit{Stop} element. Arrows are used to show precendence:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\node (start) [goal] {Start};
		\node (process) [process, below=of start] {Process};
		\node (end) [goal, below=of process] {End};
		
		\path [line] (start) -- (process);
		\path [line] (process) -- (end);
	\end{tikzpicture}
\end{figure}

Arrows can also be labelled for \textit{Decision} elements:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\node (start) [goal] {Start};
		\node (decision) [decision, below=of start] {Continue?};
		\node (process1) [process, right=of decision] {Process 1};
		\node (process2) [process, below=of process1] {Process 2};
		\node (stop) [goal, left=of decision] {Stop};

		\path [line] (start) -- (decision);
		\path [line] (decision) -- node {Yes} (process1);
		\path [line] (decision) -- node {No} (stop);
		\path [line] (process1) -- (process2);
		\path [line] (process2) -| (decision);
	\end{tikzpicture}
\end{figure}

As shown above, loops can also be created by linking a \textit{Process}
element back to a \textit{Decision} element.

The above can be adjusted to use the \textit{Function} element:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\node (start) [goal] {Start};
		\node (decision) [decision, below=of start] {Continue?};
		\node (function) [process, below right=of decision]
		{\underline{Function}};
		\node (stop) [goal, left=of decision] {Stop};
		\node (process2) [process, right=of function] {Process 2};
		\node (process1) [process, above=of process2] {Process 1};
		\node (functionbody) [process, above=of process1]
		{\underline{Function}};

		\path [line] (start) -- (decision);
		\path [line] (decision) -| node {Yes} (function);
		\path [line] (decision) -- node {No} (stop);
		\path [line] (function) -| (decision);
		\path [line] (functionbody) -- (process1);
		\path [line] (process1) -- (process2);
	\end{tikzpicture}
\end{figure}

\subsection{Pseudocode}

Pseudocode: a "made up" language meant as a middle point between actual
programming languages and natural language. Uses the following keywords:

\begin{itemize}
	\item{\texttt{START} and \texttt{STOP}: start and stop.}
	\item{\texttt{While} and \texttt{For}: loops.}
	\item{\texttt{If}, \texttt{Else if} and \texttt{Else}: conditionals.}
	\item{\texttt{Function}: function statements.}
\end{itemize}

Also uses the following rules:

\begin{itemize}
	\item{Control statements are followed by a colon.}
	\item{Blocks of code that are part of a control statement are
		indented.}
	\item{Function calls are underlined.}
	\item{\texttt{\textit{\#}} is used for comments.}
\end{itemize}

The following code snippet demonstrates an algorithm that might be used in a
bar to prepare drinks:

\begin{lstlisting}[language=pseudo]
Function(Make drink):
	If drink is martini:
		Make martini.
	Else if drink is beer:
		Get beer from fridge.
Function(Main):
	While there is another customer:
		Ask customer what drink they would like.
		%\underline{Make drink}% as requested.
		Give customer their requested drink.
		Charge customer for their drink.

START
%\underline{Main}%
STOP
\end{lstlisting}

\subsection{Python}

Python: a high level, general purpose, interpreted, dynamically typed
programming language.

High level language: abstracted away from lower level languages (C/C++,
Assembly, and machine code). The layers of abstraction used depends on the
specific function that is in the program.

General purpose language: programming language which can be used for a
variety of purposes and situations.

Interpreted language: programming language which is run line by line. Each
line is interepreted by an interepreter, which is required to run the
language. See also compiled language.

Compiled language: programming language which must be compiled before it can
be run. Requires a compiler to convert the source code to machine code. See
also interpreted language.

Statically typed language: language where the variable type must be specified
in the source code.

Dynamically typed language: language where the variable type is checked during
runtime rather than when it is compiled.

Python can be run in 2 different modes:

\begin{itemize}
	\item{Interactive mode: the user is provided with a prompt (usually
		\texttt{>>>}) to input statements. These statements are
		evaluated and the results are printed to the screen. Useful
		for testing Python features and viewing documentation.}
	\item{Script mode: the user provides Python with a script that is then
		run by Python. The interpreter runs through the script,
		interpreting each line and performing relevant actions. Useful
		for running actual code.}
\end{itemize}

\chapter{Programs and variables}

\section{Notes}

\subsection{Variables}

Variable: atomic piece of data that is used in a program. Each variable has a
unique identifier and a type.

Identifier: name of a variable. Should relate to what the variable is (for
example, a variable which stores the length of an object might have identifier
\texttt{length}.)

Type: type of data. Examples include integer, float, string, character, and
many others. A variables type dictates how operations that are performed on
that variable behave.

In Python, variables are created when you assign a value to an identifier:

\begin{lstlisting}[language=Python]
identifier = expression
\end{lstlisting}

The expression on the right of is evaluated, and that value is assigned to
the identifier. The variable has now been created, and can be used in other
parts of the program:

\begin{lstlisting}[language=Python]
>>> a = 5
>>> print(a)
5
>>> print(a + 2)
7
\end{lstlisting}

If the value of the variable needs to be changed, it must be reassigned using
the same notation as above:

\begin{lstlisting}[language=Python]
>>> b = 9
>>> print(b - 4)
5
>>> b = b - 5
>>> b
4
\end{lstlisting}

It is important to note that the equals symbol in Python is the assignment
operator.

Literal: piece of data that doesn't need to be evaluated. In the code
above, \texttt{a = 5} and \texttt{b = 9} are examples of literals being
assigned to variables.

Expression: computable product that can be evaluated. In the code above,
\texttt{b = b - 5} is an example of an expression - \texttt{b - 5} has to be
evaluated by the interpreter, and then the result is assigned to \texttt{b}.

\subsection{Types}

Variables can have different types. The built-in Python types include:

\begin{itemize}
	\item{\texttt{bool}: \texttt{True} or \texttt{False}.}
	\item{\texttt{int}: integer numbers (such as \texttt{1, -2, 4096}).}
	\item{\texttt{float}: floating point numbers (such as \texttt{10.25,
		-7.0, 3.141592653589793}).}
	\item{\texttt{str}: strings (such as \texttt{"Hello", "world",
		"Goodbye 'world'"}).}
\end{itemize}

There are more, however these are the main types that are used in everday
programming. Different types have different operations that can be performed
on them, for example:

\begin{itemize}
	\item{\texttt{2.5 + 5} evaluates to \texttt{7.5}, while \texttt{"2.5"
		+ "five"} evaluates to \texttt{"2.5five"}.}
	\item{\texttt{2.5 * 3} evaluates to \texttt{7.5}, while
		\texttt{"Hello" * 3} evaluates to \texttt{"HelloHelloHello"}.}
\end{itemize}

To find the different operations that can be performed on specific types, use
the Python builtin \texttt{help()}. For example, to find information about the
\texttt{str} type:

\begin{lstlisting}[language=Python]
>>> help(str)
\end{lstlisting}

This will output the documentation for the string type.

\end{document}
