\documentclass[a4paper,oneside]{report}

\renewcommand{\chaptername}{Week}
\setcounter{tocdepth}{3}

\usepackage[margin=2.5cm]{geometry}
\usepackage{parskip}

\usepackage{gensymb}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{listings}
\lstset{
	basicstyle=\ttfamily,
	numbers=left,
	escapechar=\%,
}

\usepackage{tikz}
\usetikzlibrary{arrows, positioning, shapes}
\tikzstyle{goal} = [draw, ellipse, minimum height=1cm, minimum width=2cm]
\tikzstyle{process} = [draw, rectangle, rounded corners, minimum height=1cm,
	minimum width=1cm]
\tikzstyle{decision} = [draw, diamond, minimum height=1cm, minimum width=2cm]
\tikzstyle{line} = [draw, -latex']

\usepackage{pgfplots}

\begin{document}

\begin{titlepage}
	\raggedleft
	\vspace*{1cm}
	\Huge{Computational Mathematics}\\
	\vspace{0.5cm}
	\Large{ECM1416}
\end{titlepage}

\null
\thispagestyle{empty}
\addtocounter{page}{-2}
\newpage

\tableofcontents

\chapter{Vectors and Introduction to Matrices}

\section{Notes}

\subsection{Asynchronous Videos}

\subsubsection{Real Coordinate space}

We have previously seen that graphs can be formed using a set of axis, for
example the 2 dimensional coordinate system requires a pair of perpendicular
axis which share a single unit of length across them. Typically the horizontal
axis is labelled \( x \) and the vertical axis \( y \). The point where these
axis meet is called the origin and is labelled \( O \). All other coordinates
are defined by an ordered pair of numbers, a tuple whose first number defines
how far to move along the \( x \)-axis and the second how far to move along
the \( y \)-axis. You can think of the axis as being a pair of number lines,
with the origin having coordinates (0, 0) and all other coordinates following
from that. The following graph displays some coordinates:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=center, axis x line=center,
			xmin=-5.5, xmax=5.5, ymin=-5.5, ymax=5.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[mark=*] coordinates {(0, 0)};
			\addplot[mark=*] coordinates {(3, 3)};
			\addplot[mark=*] coordinates {(-3, 4)};
			\addplot[mark=*] coordinates {(-3, -2)};
			\addplot[mark=*] coordinates {(4, -2)};
		\end{axis}
	\end{tikzpicture}

	The points \((0, 0), (3, 3), (-3, 4), (-3, -2), \text{ and }
	(4, -2)\) are shown here.
\end{figure}

This coordinate system can be defined more thoroughly to display coordinates
whose components are real numbers, creating a real coordinate space. The above
is an example of the real coordinate space of dimension 2, also known as
\( \mathbb{R}^2 \). A real coordinate space of dimension \( n \) is written as
\( \mathbb{R}^n \), and the coordinates on this space are an ordered tuple
are given by \( (x_1, x_2, x_3, \dots, x_n) \).

\subsubsection{Introduction to Vectors}

Let the coordinate \( P = (7, -4) \) and the coordinate \( Q = (-3, 6) \). We
can find a vector \( \overrightarrow{PQ} \) by doing \( P - Q \) element-wise,
that is \( 7 - -3 = 10 \) and \( -4 - 6 = -10 \). The vector
\( \overrightarrow{PQ} = (10, -10) \) where the first number defines how far
to move along the \( x \)-axis starting from the \( x \)-coordinate of \( P \)
and the second how far to move along the \( y \)-axis from the
\( y \)-coordinate of \( P \). This is similar to the syntax used for
coordinates, however unlike coordinates this is not limited to the axis. You
can think of a vector as being instructions of how to move between a
particular set of points, or alternatively how to move in a well defined way
from an arbritrary point. Both kinds of movement are important in linear
algebra. Some important vectors are:

\begin{itemize}
	\item{Unit vectors: these are vectors whose norm is equivalent to
		1 of the unit that is used by the axis on which they lie. For
		a given vector \( \overrightarrow{u} \) its unit vector is
		written as \( \hat{u} \). Provided that the original vector is
		not the zero vector, this unit vector will point in the same
		direction as the original vector.}
	\item{Basis vectors: in \( \mathbb{R}^2 \), these are typically (1, 0)
		and (0, 1). You may think of these as movements along the
		\( x \)- and \( y \)-axis respectively, and they form the
		basis for all the other vectors in \( \mathbb{R}^2 \). They
		don't have to be these vectors however; any 2 vectors \( a \)
		and \( b \) can form the basis vectors of \( \mathbb{R}^2 \)
		provided they are not paralell or zero vectors. All the other
		vectors in \( \mathbb{R}^2 \) can then be defined as multiples
		of these vectors and so every point in \( \mathbb{R}^2 \) can
		be reached. These may also be referred to as the standard
		unit vectors.}
	\item{Zero vectors: these are vectors whose norm is 0, and so all its
		components are also 0. For example:
		
		\[ \overrightarrow{v} = \begin{pmatrix}
			0 \\
			0 \\
		\end{pmatrix} \]

		Is an example of a zero vector.}
	\item{\( \overrightarrow{OA} \): recall that \( O \) is used to refer
		to the origin of a given coordinate space. Let \( A \) be a
		coordinate in the coordinate space. The vector
		\( \overrightarrow{OA} \) therefore refers to the vector that
		moves from the origin to the point \( A \).}
\end{itemize}

Generally speaking, vectors are often referred to with an arrow overhead,
for example \( \overrightarrow{a} = (1, 3, -4) \). The components/elements of
a vector are referred to using subscripts, for example \( a_2 = 3\).

\subsubsection{Unit Vectors and Norms}

Recall unit vectors, these are vectors whose norm is 1. The norm of a vector
refers to its length/magnitude. To find the norm of a vector
\( \overrightarrow{v} \), we do the following:

\[ |\overrightarrow{v}| = \sqrt{v_1^2 + v_2^2 + \dots + v_n^2} \]

For example, the norm of (3, 4) is \( \sqrt{3^2 + 4^2} = 5 \). We can use this
result to find the unit vector by doing the following:

\[ \hat{u} = \frac{\overrightarrow{u}}{|\overrightarrow{u}|} \]

Using the example from above, we get the unit vector \( (\frac{3}{5},
\frac{4}{5}) \). We can confirm this is in fact a unit vector by finding its
norm:

\[ \sqrt{ \frac{3}{5}^2 + \frac{4}{5}^2 } = 1 \]

Technically, there are many kinds of norm. This specific kind is called the
Euclidian norm, and generally if a norm is mentioned without some prefix then
this is the norm that is being referred to.

Some important properties of a norm:

\begin{itemize}
	\item{The norm of a vector \( \overrightarrow{v} \geq 0 \). This
		should be obvious from the fact that we square a set of values
		and then take the square root of these values.}
	\item{\( |k\overrightarrow{v}| = |k||\overrightarrow{v}| \) where
		\( k \) is some constant scalar quantity.}
	\item{\( |\overrightarrow{u} + \overrightarrow{v}| \leq
		|\overrightarrow{u}| + |\overrightarrow{v}| \): this is
		the triangle inequality. Essentially, one side of a triangle
		will always be shorter than or equal to the other 2 sides
		added together.}
	\item{\( |\overrightarrow{u} - \overrightarrow{v}| \geq
		|\overrightarrow{u}| - |\overrightarrow{v}| \) this is
		the reverse triangle inequality.}
\end{itemize}

The norm of a vector may have many different equivalent names, including
length and magnitude.

Before we move on, we should talk about the standard unit vectors. Recall
basis vectors: these are vectors with which we can describe movement in the
entire real coordinate space for that particular dimension. The standard unit
vectors (also known as canonical vectors) are a set of unit basis vectors that
are used to describe movement in the direction of each of the axis. For
example, in \( \mathbb{R}^2 \), the standard unit vectors are: \( \hat{i} =
(1, 0) \) for movement in the \( x \) direction and \( \hat{j} = (0, 1) \) for
movement in the \( y \) direction. The vector \( \overrightarrow{v} =
(a, b) \) can therefore be described as \( \overrightarrow{v} = a\hat{i} +
b\hat{j} \). This principle can be expanded to higher dimensions, for example
in \( \mathbb{R}^3 \) we have \( \hat{i} = (1, 0, 0) \), \( \hat{j} =
(0, 1, 0) \) and \( \hat{k} = (0, 0, 1) \) for movement in the \( x \), \( y \)
and \( z \) directions.

\subsubsection{Vector Arithmetic}

Vectors can be scaled, that is we can increase and decrease the magnitude of a
vector by multiplying a vector by a scalar. We do this by multiply each
component of the vector by the given scalar. This is done as follows:

\[ \lambda \times \begin{pmatrix}
	v_1 \\
	v_2 \\
	\vdots \\
	v_n \\
\end{pmatrix} = \begin{pmatrix}
	\lambda v_1 \\
	\lambda v_2 \\
	\vdots \\
	\lambda v_n \\
\end{pmatrix} \]

For example:

\[ 2 \times \begin{pmatrix}
	4 \\
	6 \\
\end{pmatrix} = \begin{pmatrix}
	8 \\
	12 \\
\end{pmatrix} \]

\[ -1 \times \begin{pmatrix}
	4 \\
	6 \\
\end{pmatrix} = \begin{pmatrix}
	-4 \\
	-6 \\
\end{pmatrix} \]

\[ \frac{1}{2} \times \begin{pmatrix}
	4 \\
	6 \\
\end{pmatrix} = \begin{pmatrix}
	2 \\
	3 \\
\end{pmatrix} \]

The graphical representation of the above is as follows:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=center, axis x line=center,
			xmin=-15.5, xmax=15.5, ymin=-15.5, ymax=15.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=4, v=6}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=8, v=12}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=-4, v=-6}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=2, v=3}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Scaled vectors
\end{figure}

As can be seen, there are 6 cases for some given scalar \( \lambda \):

\begin{itemize}
	\item{\( \lambda > 1 \): the vectors magnitude will be increased (it
		gets bigger).}
	\item{\( \lambda = 1 \): the vector is unchanged.}
	\item{\( 0 < \lambda < 1 \): the vectors magnitude will be decreased
		(it shrinks).}
	\item{\( -1 < \lambda < 0 \): the vectors direction is reversed and
		its magnitude decreased (it flips and shrinks).}
	\item{\( \lambda = -1 \): the vectors direction is reversed.}
	\item{\( \lambda < -1\): the vectors direction is reversed and
		its magnitude increased (it flips and gets bigger).}
\end{itemize}

Besides scaling vectors, we can also add and subtract vectors. We do this by
adding or subtracting component-wise. This is done as follows:

\[ \begin{pmatrix}
	v_1 \\
	v_2 \\
	\vdots \\
	v_n \\
\end{pmatrix} + \begin{pmatrix}
	u_1 \\
	u_2 \\
	\vdots \\
	u_n \\
\end{pmatrix} = \begin{pmatrix}
	v_1 + u_1 \\
	v_2 + u_2 \\
	\vdots \\
	v_n + u_n \\
\end{pmatrix} \]

For example:

\[ \begin{pmatrix}
	1 \\
	-3 \\
\end{pmatrix} +
\begin{pmatrix}
	-2 \\
	-4 \\
\end{pmatrix} = \begin{pmatrix}
	-1 \\
	-7 \\
\end{pmatrix} \]

\[ \begin{pmatrix}
	1 \\
	2 \\
\end{pmatrix} -
\begin{pmatrix}
	6 \\
	-3 \\
\end{pmatrix} = \begin{pmatrix}
	-5 \\
	5 \\
\end{pmatrix} \]

The graphical representation of the above is as follows:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=center, axis x line=center,
			xmin=-10.5, xmax=10.5, ymin=-10.5, ymax=10.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=1, v=-3}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=-2, v=-4}, -latex']
			coordinates {(1, -3)};
			\addplot[quiver={u=-1, v=-7}, -latex']
			coordinates {(0, 0)};

			\addplot[quiver={u=1, v=2}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=-6, v=3}, -latex']
			coordinates {(1, 2)};
			\addplot[quiver={u=-5, v=5}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Sum of vectors
\end{figure}

Notice that in the case of the second example, the direction of the second
vector has been reversed (as per the discussion about scaling vectors
previously). Also note that 2 vectors can only be added or subtracted if they
are conformable, that is both vectors are of the same dimension. You cannot
add a vector of dimension 3 to one of dimension 2. One final note: vector
addition is commutative, that is:

\[ \overrightarrow{u} + \overrightarrow{v} = \overrightarrow{v} +
\overrightarrow{u} \]

This can be demonstrated by swapping the order in which the examples above
are laid out - you should find that you still end up at the same vector.
Vector addition is also associative, that is:

\[ (\overrightarrow{u} + \overrightarrow{v}) + \overrightarrow{w} =
\overrightarrow{u} + (\overrightarrow{v} + \overrightarrow{w}) \]

This can be demonstrated in a similar manner to the above: if you were
to graph this with any 3 vectors you should find that it doesn't matter
in which order you add the vectors, you will always end up at the same
vector.

There are other vector operations, namely the scalar product and the vector
product. These will be covered seperately.

\subsubsection{Scalar Product}

The scalar product (also known as the dot product because the \( \cdot \)
character is used to denote it) has two mathematical definitions. The first
is as follows:

\[ \overrightarrow{u} \cdot \overrightarrow{v} = \displaystyle\sum_{i=1}^{n}
u_i v_i \]

Where \( \overrightarrow{u} \) and \( \overrightarrow{v} \) are the vectors
for which we are taking the dot product and \( n \) is the dimension of these
vectors (note that both vectors must have the same dimension in order for this
operation to work). This is known as the algebraic definition. For example,
to find the dot product of \( \overrightarrow{u} = (1, -4, 3) \) and
\( \overrightarrow{v} = (-7, 8, 0) \), we do:

\[ \begin{pmatrix}
	1 \\
	-4 \\
	3 \\
\end{pmatrix} \cdot \begin{pmatrix}
	-7 \\
	8 \\
	0 \\
\end{pmatrix} = (1 \times -7) + (-4 \times 8) + (3 \times 0) = -39 \]


Before we describe the second definition (known as the geometric definition),
let us first consider 2 vectors whose tails are located on the same point:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=center, axis x line=center,
			xmin=-8.5, xmax=8.5, ymin=-8.5, ymax=8.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=3, v=4}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=-6, v=-3}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	2 vectors in space
\end{figure}

Consider the angle between these 2 vectors, we can denote this angle by
\( \theta \). Imagine taking one of the vectors and creating a number line
from it whose 0 starts at the tail of the vector and increases positively
in the direction the vector is pointing. Then, take the other vector and
project it onto this new line. Finally, multiply the lengths of the vectors
along this line by each other to find their scalar product.

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=center, axis x line=center,
			xmin=-8.5, xmax=8.5, ymin=-8.5, ymax=8.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=3, v=4}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=-6, v=-3}, -latex']
			coordinates {(0, 0)};

			\addplot[mark=none]{4/3 * x};
			\addplot[quiver={u=-2.4,v=1.8}, dashed]
			coordinates {(-3.6, -4.8)};
			\addplot[quiver={u=-3.6, v=-4.8}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Dot product
\end{figure}

In the example above, we had 2 vectors: \( \overrightarrow{u} = (3, 4) \) and
\( \overrightarrow{v} = (-6, -3) \). We can choose either of these vectors, it
doesn't make a difference. \( \overrightarrow{u} \) has length 5, and we
project \( \overrightarrow{v} \) onto the imaginary line that we created. The
length of this line can be calculated from either the coordinates or you can
use \( | \overrightarrow{v} | \cos{\theta} \), which gives length 6. As the
projection of \( \overrightarrow{v} \) is pointing in the opposite direction
of \( \overrightarrow{v} \), we do \( 5 \times -6 \) to get a dot product of
\( -30 \). The second option gives rise to the second definition of the dot
product:

\[ \overrightarrow{u} \cdot \overrightarrow{v} =
| \overrightarrow{u} | | \overrightarrow{v} | \cos{\theta} \]

We can use this result to find the angle between 2 vectors, for example to
find the angle between \( \overrightarrow{u} = (-3, 4, -5) \) and
\( \overrightarrow{v} = (8, 1, -5) \):

\[ \overrightarrow{u} \cdot \overrightarrow{v} = (-3 \times 8) + (4 \times 1)
+ (-5 \times -5) = 5 \]

\[ | \overrightarrow{u} | | \overrightarrow{v} | = \sqrt{50} \times \sqrt{90}
= \sqrt{4500} = 30 \sqrt{5} \]

\[ \cos{\theta} = \frac{\overrightarrow{u} \cdot \overrightarrow{v}}
{| \overrightarrow{u} | | \overrightarrow{v} |} = \frac{5}{30 \sqrt{5}} =
\frac{\sqrt{5}}{30} \]

\[ \theta = \arccos{\frac{\sqrt{5}}{30}} \approx 85.73 \degree \]

There are some important consequences of this:

\begin{itemize}
	\item{\( \overrightarrow{u} \cdot \overrightarrow{v} > 0 \): both
		vectors are pointing in the same direction.}
	\item{\( \overrightarrow{u} \cdot \overrightarrow{v} = 0 \): both
		vectors are perpendicular.}
	\item{\( \overrightarrow{u} \cdot \overrightarrow{v} < 0 \): both
		vectors are pointing in opposite directions.}
\end{itemize}

This is particularly useful in coordinate spaces greater than
\( \mathbb{R}^2 \) where it may not be obvious if 2 vectors are pointing in
the same direction or even what the angle between 2 vectors actually is.

The dot product is commutative: that is:

\[ \overrightarrow{u} \cdot \overrightarrow{v} = \overrightarrow{v} \cdot
\overrightarrow{v} \]

This should be obvious from both definitions of the dot product, given
that we use multiplication and addition in them (which are both associative).
The dot product, however, is not associative, as the dot product returns a
scalar, and the dot product between a scalar and a vector is not defined.

It should be emphasised that the result of the dot product is a scalar, that
is, taking the scalar product of 2 vectors will result in a number, not a
vector. This is the reason the dot product is synonymously called the scalar
product.

\subsubsection{Transformations with Vectors}

The last topic to cover related purely to vectors are transformations to do
with vectors. There are 3 main transformations:

\begin{itemize}
	\item{Translation: moving a vector from some position \( P \) to a
		new position \( P' \).}
	\item{Scaling: scaling a vector around the origin with some scale
		factor \( \lambda \).}
	\item{Rotation: rotating the direction of a vector about the origin
		by some angle \( \theta \).}
\end{itemize}

Recall vector arithmetic, in particular vector addition and vector
multiplication by a scalar. These are, in essence, the algebraic perspective
of vector translation and vector scaling (respectively). As they have already
been covered previously, rotation is given the focus here.

When rotating a vector, we typically place its tail at the origin and rotate
it anticlockwise by some angle \( \theta \) with respect to the positive
\( x \)-axis (note in this context we assume that the vectors are in
\( \mathbb{R}^2 \), there are other solutions for higher dimensions but they
are not covered here). Assuming we have some vector \( \overrightarrow{v} \),
the resulting vector \( \overrightarrow{v}' \) is given by the following:

\[ \overrightarrow{v}' = \begin{pmatrix}
	v_x \cos{\theta} - v_y \sin{\theta} \\
	v_x \sin{\theta} + v_y \cos{\theta} \\
\end{pmatrix} \]

Where \( v_x \) and \( v_y \) are the \( x \) and \( y \) components of
\( \overrightarrow{v} \). For example, to rotate the vector
\( \overrightarrow{v} = (3, 4) \) by 30\degree, we do:

\[ \overrightarrow{v}' = \begin{pmatrix}
	3 \cos{30\degree} - 4 \sin{30\degree} \\
	3 \sin{30\degree} + 4 \cos{30\degree} \\
\end{pmatrix} = \begin{pmatrix}
	\frac{3 \sqrt{2}}{2} - 2 \\
	\frac{3}{2} + 2 \sqrt{3} \\
\end{pmatrix} = \begin{pmatrix}
	\frac{-4 + 3 \sqrt{3}}{2} \\
	\frac{3 + 4 \sqrt{3}}{2} \\
\end{pmatrix} \]

The geometric interpretation of the above is as follows:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=left, axis x line=left,
			xmin=0, xmax=6.5, ymin=0, ymax=6.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=3, v=4}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=0.598, v=4.964}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Rotation with vectors
\end{figure}

One final situation and problem to consider is rotation about an arbitrary
point. Fortunately the solution is relatively simple. Let's say that we
want to rotate vector \( \overrightarrow{v} \) around some point \( P \) by
some angle \( \theta \). We create a new vector \( \overrightarrow{t} =
\overrightarrow{OP} \) and do the following:

\[ \overrightarrow{v_t} = \overrightarrow{v} - \overrightarrow{t} \]

We then perform the rotation. After the rotation is performed, we do:

\[ \overrightarrow{v}' = \overrightarrow{v_{r,t}} + \overrightarrow{t} \]

Essentially, we are translating the entire coordinate space to shift the
origin to \( P \). We then perform the rotation, and afterwards shift the
entire coordinate space back to where it should be. This is method can
also be used for scaling about an angle, simply replace the operation of
rotation with the operation of vector multiplication with a scalar.

\subsubsection{Introduction to Matrices}

A matrix is a rectangular array of elements that are arranged in table-like
structure. A matrix has a fixed number of rows and columns, and each cell
in the matrix is occupied by a number, expression or variable. For example:

\[ A_{m,n} = \begin{pmatrix}
	a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
	a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	a_{m,1} & a_{m,2} & \cdots & a_{m,n}
\end{pmatrix} \]

The above matrix is known as an \( m \times n \) matrix: it has \( m \) rows
and \( n \) columns. Each element is referred to by a subscript with the first
part of the subscript referring to its row and the second part referring to
its column. Some examples of matrices are as follows:

\[ A_{2, 3} = \begin{pmatrix}
	2 & 5 & -7 \\
	3 & 0 & 1 \\
\end{pmatrix}, B_{2, 2} = \begin{pmatrix}
	1 & 0 \\
	0 & 2 \\
\end{pmatrix}, C_{2, 1} = \begin{pmatrix}
	3 \\
	4 \\
\end{pmatrix} \]

Some important kinds of matrices are described below:

\begin{itemize}
	\item{Square matrices: these are matrices where the number of rows
		is equal to the number of columns, that is \( m = n \). These
		are particularly important in linear transformations. They
		are also the only kinds of matrices which are conformable for
		inversion and exponentiation.}
	\item{Identity matrices: these are square matrices where the diagonal
		from the top left to the bottom right consists of 1s and all
		other elements consists of 0s. For example:
		
		\[ I_2 = \begin{pmatrix}
			1 & 0 \\
			0 & 1 \\
		\end{pmatrix} \]

		These are useful in matrix multiplication as they leave
		matrices completely unchanged. and can be thought of as the
		number 1 for matrices. They are often denoted with the letter
		\( I \) and sometimes have a subscript denoting their size.}
	\item{Zero matrices: these are matrices where all the elements in the
		matrix are 0. Similar to the identity matrix they can be
		thought of as the number 0 for matrices, however unlike the
		identity matrix they are not necessarily square. They are
		typically denoted with the letter \( Z \) and may have a
		subscript denoting their rows and columns.}
	\item{Vectors: you can think of a vector as being a special case of
		a matrix where the number of columns is limited to 1 and the
		number of rows is the dimension of the vector. This mindset
		is useful when applying matrices to vectors which is common
		in linear algebra.}
\end{itemize}

There are 3 main unary operations that can occur on a single matrix:

\begin{itemize}
	\item{Transposing a matrix.}
	\item{Finding the inverse of a matrix such that \( AA^{-1} = I \)
		(applies to square matrices only).}
	\item{Finding the determinant of a matrix (applies to square matrices
		only).}
\end{itemize}

The second 2 will be discussed at a later time, however it is important to
consider the transpose of a matrix as it is very useful for linear
transformations and plays an important role in the latter 2 operations.

The transpose of a given matrix is typically denoted with a superscripted
\( T \), for example the transpose of some matrix \( A \) is \( A^T \). To find
the transpose of a matrix, we perform the following:

\begin{enumerate}
	\item{Write the rows of \( A \) as the columns of \( A^T \).}
	\item{Write the columns of \( A \) as the rows of \( A^T \).}
\end{enumerate}

For example:

\[ A = \begin{pmatrix}
	1 & 2 \\
	3 & 4 \\
\end{pmatrix}, A^T = \begin{pmatrix}
	1 & 3 \\
	2 & 4 \\
\end{pmatrix} \]

Transposes can also be applied to non-square matrices, for example:

\[ B = \begin{pmatrix}
	2 & 3 & 5 \\
	-1 & 4 & 9 \\
\end{pmatrix}, B^T = \begin{pmatrix}
	2 & -1 \\
	3 & 4 \\
	5 & 9 \\
\end{pmatrix} \]

This operation will come in handy later so it is important to remember it.

\subsubsection{Matrix Arithmetic}

Matrices can be scaled, that is it is possible to scale up a matrix by
multiplying a matrix by a scalar. We do this by multiplying all the elements
in the matrix by the given scalar. This is done as follows:

\[ \lambda \times \begin{pmatrix}
	a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
	a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	a_{m,1} & a_{m,2} & \cdots & a_{m,n}
\end{pmatrix} = \begin{pmatrix}
	\lambda a_{1,1} & \lambda a_{1,2} & \cdots & \lambda a_{1,n} \\
	\lambda a_{2,1} & \lambda a_{2,2} & \cdots & \lambda a_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	\lambda a_{m,1} & \lambda a_{m,2} & \cdots & \lambda a_{m,n}
\end{pmatrix} \]

For example:

\[ 3 \times \begin{pmatrix}
	-1 & 3 & 5 \\
	2 & -4 & 8 \\
\end{pmatrix} = \begin{pmatrix}
	-3 & 0 & 15 \\
	6 & -12 & 24 \\
\end{pmatrix} \]

Matrices can also be added to each other. The sum of 2 matrices only exists
if both matrices are of the same dimension, that is, both matrices have the
same number of rows and the same number of columns. The sum of 2 matrices can
be found as follows:

\[ \begin{pmatrix}
	a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
	a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	a_{m,1} & a_{m,2} & \cdots & a_{m,n}
\end{pmatrix} + \begin{pmatrix}
	b_{1,1} & b_{1,2} & \cdots & b_{1,n} \\
	b_{2,1} & b_{2,2} & \cdots & b_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	b_{m,1} & b_{m,2} & \cdots & b_{m,n}
\end{pmatrix} = \begin{pmatrix}
	(a + b)_{1,1} & (a + b)_{1,2} & \cdots & (a + b)_{1,n} \\
	(a + b)_{2,1} & (a + b)_{2,2} & \cdots & (a + b)_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	(a + b)_{m,1} & (a + b)_{m,2} & \cdots & (a + b)_{m,n}
\end{pmatrix} \]

For example:

\[ \begin{pmatrix}
	2 & 0 \\
	3 & 4 \\
\end{pmatrix} + \begin{pmatrix}
	-1 & 3 \\
	4 & 4 \\
\end{pmatrix} = \begin{pmatrix}
	1 & 3 \\
	7 & 8 \\
\end{pmatrix} \]

To subtract to matrices, we simply multiply the second matrix by the scalar
\( -1 \) and then find the sum of the two matrices. For example:

\[ \begin{pmatrix}
	7 & 3 \\
	1 & 0 \\
\end{pmatrix} - \begin{pmatrix}
	4 & 3 \\
	-1 & 5 \\
\end{pmatrix} = \begin{pmatrix}
	7 & 3 \\
	1 & 0 \\
\end{pmatrix} + \begin{pmatrix}
	-4 & -3 \\
	1 & -5 \\
\end{pmatrix} = \begin{pmatrix}
	3 & 0 \\
	2 & -5 \\
\end{pmatrix} \]

Matrix addition is commutative, that is:

\[ A + B = B + A \]

Matrix addition is also associative, that is:

\[ (A + B) + C = A + (B + C) \]

These should be obvious from the underlying operation of addition also being
commutative and associative, and so matrix addition must also be commutative
and associative.

The last matrix arithmetic operation to be covered is matrix multiplication.
Given some matrix \( A \) with dimensions \( m \times n \) and some matrix
\( B \) with dimensions \( n \times p \) the matrix product \( C = AB \).
This new matrix has dimensions \( m \times p \). This is done as follows:

\[ A = \begin{pmatrix}
	a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
	a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	a_{m,1} & a_{m,2} & \cdots & a_{m,n}
\end{pmatrix}, B = \begin{pmatrix}
	b_{1,1} & b_{1,2} & \cdots & b_{1,n} \\
	b_{2,1} & b_{2,2} & \cdots & b_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	b_{m,1} & b_{m,2} & \cdots & b_{m,n} \\
\end{pmatrix} \]

\[ C = \begin{pmatrix}
	c_{1,1} & c_{1,2} & \cdots & c_{1,n} \\
	c_{2,1} & c_{2,2} & \cdots & c_{2,n} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	c_{m,1} & c_{m,2} & \cdots & c_{m,n}
\end{pmatrix} \]

Where:

\[ c_{i,j} = a_{i,1}b{1,j} + a_{i,2}b_{2,j} + \dots + a_{i,n}b_{n,j} =
\displaystyle\sum_{k=1}^n a_{i,k}b_{k,j} \]

For \( i = 1, \dots, m \) and \( j = 1, \dots, p \). For example:

\[ \begin{pmatrix}
	2 & 3 \\
	-1 & 4 \\
\end{pmatrix} \begin{pmatrix}
	-6 & 5 \\
	9 & 8 \\
\end{pmatrix} = \begin{pmatrix}
	2 \times -6 + 3 \times 9 & 2 \times 5 + 3 \times 8 \\
	-1 \times -6 + 4 \times 9 & -1 \times 5 + 4 \times 8 \\
\end{pmatrix} = \begin{pmatrix}
	15 & 34 \\
	42 & 27 \\
\end{pmatrix} \]

\[ \begin{pmatrix}
	3 & 6 \\
	-3 & 1 \\
	4 & -2 \\
	5 & 7 \\
\end{pmatrix} \begin{pmatrix}
	1 & 0 & 8 \\
	-1 & 3 & 7 \\
\end{pmatrix} = \begin{pmatrix}
	3 \times 1 + 6 \times -1 & 3 \times 0 + 6 \times 3 & 3 \times 8 + 6 \times 7 \\
	-3 \times 1 + 1 \times -1 & -3 \times 0 + 1 \times 3 & -3 \times 8 + 1 \times 7 \\
	4 \times 1 + -2 \times -1 & 4 \times 0 + -2 \times 3 & 4 \times 8 + -2 \times 7 \\
	5 \times 1 + 7 \times -1 & 5 \times 0 + 7 \times 3 & 5 \times 8 + 7 \times 7 \\
\end{pmatrix} = \begin{pmatrix}
	-9 & 18 & 66 \\
	2 & 3 & -17 \\
	-2 & -6 & 18 \\
	-12 & 21 & 89 \\
\end{pmatrix} \]

The matrix product of two matrices only exists if the two matrices are
conformable. More specifically, if you had two matrices \( A \) and \( B \)
and you wanted to find the matrix product \( AB \), the number of columns in
\( A \) is equal to the number of rows in \( B \). If this is not the case,
then the matrix product doesn't exist. If this is the case then the resulting
matrix will have dimension \( m \times p \) (where \( m \) is the number of
columns in matrix \( A \) and \( p \) is the number of rows in matrix
\( B \)).

Matrix multiplication is associative, that is:

\[ (AB) C = A (BC) \]

It should be noted that matrix multiplication is not commutative, that is

\[ AB \neq BA \]

This is especially true when the matrices have different dimensions: if we
say \( A \) is of dimension \( m \times n \) and \( B \) is of dimension
\( p \times q \) then \( AB \) is defined with dimensions \( m \times q \)
(if and only if \( n = p \)) while \( BA \) is defined with dimensions
\( p \times n \) (if and only if \( m = q \)). In the case of differently
sized matrices (that is, \( m = q \neq n = p \)) then the resulting matrices
will have different dimensions and so cannot be equal. The only other case is
if \( m = q = n = p \), that is both matrices are square with the same
dimension. Even in this case, there are a variety of examples that show that
matrix multiplication is not commutative:

\[ A = \begin{pmatrix}
	2 & 0 \\
	0 & 0 \\
\end{pmatrix}, B = \begin{pmatrix}
	0 & 3 \\
	0 & 0 \\
\end{pmatrix} \]

\[ AB = \begin{pmatrix}
	0 & 6 \\
	0 & 0 \\
\end{pmatrix}, BA = \begin{pmatrix}
	0 & 0 \\
	0 & 0 \\
\end{pmatrix} \]

\subsubsection{Transformations with Matrices}

Recall vectors, and that they are a special case of matrices where the number
of columns is limited to 1 and the number of rows is the dimension of the
matrix. Consider the following:

\[ A = \begin{pmatrix}
	1 & 2 \\
	-1 & 2 \\
\end{pmatrix}, \overrightarrow{v} = \begin{pmatrix}
	2 \\
	3 \\
\end{pmatrix} \]

\[ \overrightarrow{v}' = A\overrightarrow{v} = \begin{pmatrix}
	8 \\
	4 \\
\end{pmatrix} \]

Consider the geometric interpretation of the above by plotting both
\( \overrightarrow{v} \) and \( \overrightarrow{v}' \):

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=left, axis x line=left,
			xmin=0, xmax=10.5, ymin=0, ymax=6.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=2, v=3}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=8, v=4}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Transformations with matrices
\end{figure}

As can be seen, we have applied the matrix \( A \) to the vector
\( \overrightarrow{v} \) to create a new vector \( \overrightarrow{v}' \). You
can therefore consider matrix multiplication between a matrix and a vector as
applying that matrix to the vector to transform it into a new vector. It is
possible to describe several different kinds of transformations using matrices
including:

\begin{itemize}
	\item{Scaling}
	\item{Rotation}
	\item{Shearing}
\end{itemize}

We do not describe translation using matrices as vectors are more convenient
for this task.

We already know that we can scale a vector by multiplying it by a scalar. The
matrix equivalent of this is to multiply the identity matrix by a scalar and
then multiply that new matrix by the vector you wish to transform. This is
done as follows:

\[ \begin{pmatrix}
	\lambda & 0 \\
	0 & \lambda \\
\end{pmatrix} \begin{pmatrix}
	v_x \\
	v_y \\
\end{pmatrix} = \begin{pmatrix}
	\lambda v_x \\
	\lambda v_y \\
\end{pmatrix} \]

The value of \( \lambda \) determines the scale factor of the transformation.
Refer to the list shown earlier to determine the corresponding geometrical
effect given some value of \( \lambda \).

This transformation can be adjusted to only scale along a specific axis, or
alternatively to scale differently along different axis. This is done as
follows:

\[ \begin{pmatrix}
	\lambda_x & 0 \\
	0 & \lambda_y \\
\end{pmatrix} \begin{pmatrix}
	v_x \\
	v_y \\
\end{pmatrix} = \begin{pmatrix}
	\lambda_x v_x \\
	\lambda_y v_y \\
\end{pmatrix} \]

Like before, the value of \( \lambda_x \) and \( \lambda_y \) are the scale
factor in the \( x \) and \( y \) directions. However if we choose to make
either of \( \lambda_x \) or \( \lambda_y \) equal to 1, then the
transformation will only occur in the other direction (provided that the other
value of \( \lambda \) is not also equal to 1). For example:

\begin{itemize}
	\item{Scaling in the \( x \) direction only:

\[ \begin{pmatrix}
	3 & 0 \\
	0 & 1 \\
\end{pmatrix} \begin{pmatrix}
	3 \\
	4 \\
\end{pmatrix} = \begin{pmatrix}
	9 \\
	4 \\
\end{pmatrix} \]
		}
	\item{Scaling in the \( y \) direction only:

\[ \begin{pmatrix}
	1 & 0 \\
	0 & -2 \\
\end{pmatrix} \begin{pmatrix}
	3 \\
	4 \\
\end{pmatrix} = \begin{pmatrix}
	3 \\
	-8 \\
\end{pmatrix} \]
		}
	\item{Scaling seperately in both the \( x \) and \( y \)
		directions:

\[ \begin{pmatrix}
	3 & 0 \\
	0 & -2 \\
\end{pmatrix} \begin{pmatrix}
	3 \\
	4 \\
\end{pmatrix} = \begin{pmatrix}
	9 \\
	-8 \\
\end{pmatrix} \]
		}
	\item{Scaling congruently in both the \( x \) and \( y \) directions:

		\[ \begin{pmatrix}
			-2 & 0 \\
			0 & -2 \\
		\end{pmatrix} \begin{pmatrix}
			3 \\
			4 \\
		\end{pmatrix} = \begin{pmatrix}
			-6 \\
			-8 \\
		\end{pmatrix} \]
		}
\end{itemize}

The geometric effects of the above are as follows:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=center, axis x line=center,
			xmin=-10.5, xmax=10.5, ymin=-10.5, ymax=10.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=3, v=4}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=9, v=4}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=3, v=-8}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=9, v=-8}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=-6, v=-8}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Scaling with matrices
\end{figure}

Recall the rotation with vectors. We can deduce the rotation matrix from this
by reverse engineering the rotation vector to give the following
transformation:

\[ \begin{pmatrix}
	\cos{\theta} & -\sin{\theta} \\
	\sin{\theta} & \cos{\theta} \\
\end{pmatrix} \begin{pmatrix}
	v_x \\
	v_y \\
\end{pmatrix} = \begin{pmatrix}
	v_x \cos{\theta} - v_y \sin{\theta} \\
	v_x \sin{\theta} + v_y \cos{\theta} \\
\end{pmatrix} \]

This transformation behaves in exactly the same way as the original: the
vector is rotated by some angle \( \theta \) anticlockwise with respect to
the positive \( x \)-axis. As this has been already described and
demonstrated earlier, it will not be demonstrated here.

Shearing is a new transformation whereby the one component or coordinate of
a vector is left unchanged while the other is increased by a multiple of the
unchanged coordinate. This can be done either paralell to the \( x \)-axis
(leaving the \( y \) coordinate unchanged) or paralell to the \( y \)-axis
(leaving the \( x \) coordinate unchanged). This is done as follows:

\begin{itemize}
	\item{Shear paralell to the \( x \)-axis:

		\[ \begin{pmatrix}
			1 & \lambda \\
			0 & 1 \\
		\end{pmatrix} \begin{pmatrix}
			v_x \\
			v_y \\
		\end{pmatrix} = \begin{pmatrix}
			v_x + \lambda v_y \\
			v_y \\
		\end{pmatrix} \]
		}
	\item{Shear paralell to the \( y \)-axis:

		\[ \begin{pmatrix}
			1 & 0 \\
			\lambda & 1 \\
		\end{pmatrix} \begin{pmatrix}
			v_x \\
			v_y \\
		\end{pmatrix} = \begin{pmatrix}
			v_x \\
			v_y + \lambda v_x \\
		\end{pmatrix} \]
		}
\end{itemize}

Shearing is often used to 

An example of a shear paralell to the \( x \)-axis is as follows:

\[ \begin{pmatrix}
	1 & 3 \\
	0 & 1 \\
\end{pmatrix} \begin{pmatrix}
	3 \\
	2 \\
\end{pmatrix} = \begin{pmatrix}
	1 \times 3 + 3 \times 2 \\
	0 \times 3 + 1 \times 2 \\
\end{pmatrix} = \begin{pmatrix}
	9 \\
	2 \\
\end{pmatrix} \]

An example of a shear paralell to the \( y \)-axis is as follows:

\[ \begin{pmatrix}
	1 & 0 \\
	2 & 1 \\
\end{pmatrix} \begin{pmatrix}
	4 \\
	2 \\
\end{pmatrix} = \begin{pmatrix}
	1 \times 4 + 0 \times 2 \\
	2 \times 4 + 1 \times 2 \\
\end{pmatrix} = \begin{pmatrix}
	4 \\
	10 \\
\end{pmatrix} \]

The geometric representation of both of the above is as follows:

\begin{figure}[h]
	\centering
	\begin{tikzpicture}
		\begin{axis}[axis y line=left, axis x line=left,
			xmin=0, xmax=10.5, ymin=0, ymax=10.5,
			xlabel=\( x \), ylabel=\( y \)]
			\addplot[quiver={u=3, v=2}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=9, v=2}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=4, v=2}, -latex']
			coordinates {(0, 0)};
			\addplot[quiver={u=4, v=10}, -latex']
			coordinates {(0, 0)};
		\end{axis}
	\end{tikzpicture}

	Shears paralell to the \( x \)- and \( y \)- axis
\end{figure}

The last thing to consider is the idea of performing multiple transformations,
for example scaling and then rotating. This can be done through matrix
multiplication: we first multiply the scaling matrix by the vector to get a
new vector, and then multiply this new vector by the rotation matrix. Matrix
multiplication has an associative property, therefore we can equivalently
make a transformation matrix by multiplying the scaling and rotation matrces
and then multiplying this new matrix by the vector to get a new vector. This
is done as follows:

\[ \overrightarrow{v}' = SR\overrightarrow{v} \]

It should be noted that while the associative property holds for matrix
multiplication, the commutative property does not. This means that the order
of transformations does matter, that is if you reverse the order of
transformations (rotate and then scale) you will not necessarily end up with
the same vector.

\subsection{Synchronous Lectures}

\section{Workshops}

Workshop exercises for this week go here.

\section{Exercises}

Solution to exercise sheets set this week go here.

\end{document}
